---
title: "Barnacle goose IPM"
author: "Kate Layton-Matthews"
date: "26 10 2023"
output:
  pdf_document: default
  html_document:
    df_print: paged
vignette: |
  %\VignetteIndexEntry{Simple IPM Example} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Integral Projection Models in PaGAn

There are a number of fantastic packages for performing analysis of data using integral projection models (IPMs) in R such as [IPMpack](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210x.12001) and [impr](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.13683).  PaGAn also has its own implementation of Integral Projection Models. The PaGAn implementation of IPM has been designed with a focus on flexibility and an ability to embed the outputs from these models into further analysis. Under the hood PaGAn uses Bayesian techniques (powered by [NIMBLE](https://www.tandfonline.com/doi/full/10.1080/10618600.2016.1172487)) which means that full posterior estimates of the coefficients in vital rate functions, kernels, and population-level summaries.

```{r libraries, results='hide', warning=FALSE, message=FALSE}
library(nimble)
library(parallel)
library(coda)
library(DHARMa)
library(ggplot2)
library(fields)

source("https://raw.githubusercontent.com/NINAnor/IPM2IPM/master/PaGAn/R/mcmcInternals.R")
source("https://raw.githubusercontent.com/NINAnor/IPM2IPM/master/PaGAn/R/glmNIMBLE.R")
source("https://raw.githubusercontent.com/NINAnor/IPM2IPM/master/PaGAn/R/ipmNIMBLE.R")
source("Scripts/megaMatrixAssembly.R")
```

## Example Dataset

This example uses individual-based data on barnacle geese (branta leucopsis) collected at their breeding grounds at Ny-Alesund, Svalbard between 1991 and 2017. See Layton-Matthews et al. (2019, Journal Animal Ecology) for details on the data collection methods: https://doi.org/10.1111/1365-2656.13001. Demographic data (exData.csv file) includes individual (N=1669) observational data, with columns describing age class (a, fledging = 0 and adult = 1), whether an individual survived from year t to t+1 (surv), body mass in grams for year t (mass) and t+1 (mass_next). Fecundity is recorded by two parameters: whether females had any offspring or not (repF, 0 or 1) and the number of fledgings per female for those that bred successfully (Nfledgling).

```{r loadData}
# Load data and show a snippet of the dataframe
exData <- read.csv("Example_BarnacleGeese/exData.csv",sep=";")
knitr::kable(head(exData))

# Save the range of body mass values for fledgings and adults seperately
za0_range <- round(range(exData$mass[exData$a==0],na.rm=T))
za1_range <- round(range(exData$mass[exData$a==1],na.rm=T))

```

## Vital Rate Regressions

The first stage for running an IPM is to run the vital rate regressions.  In PaGAn the user is allowed to have as many vital rate regressions as they need for their analysis and they can give those any name that they require. For this case study of the barnacle geese we specify four regression models; one of survival, one for growth and two sub-models for fecundity. 

We start with the survival regression function, which is formulated as two sub-models. This vital rate is a logistic regression model with the survival status as the response and mass as a predictor variable. Sub-model #1 is surival for fledlings and #2 is for adults, which are both a function of mass at year t. The data are assumed to have a binomial error distribution and a logit link function is used (NB a observation model with a sex effect needs to be included here).

```{r survSpec}
# Define the model and the error family
surv_f.modelFormula <- cbind(surv, 1 - surv) ~ mass
surv_f.errorFamily <- binomial(link = "logit")

# Sub-set the data frame so only the columns with relevant data are included
surv_f.inputData <- exData[!is.na(exData$surv) & 
                           !is.na(exData$mass) & exData$a==0, ]

# Define the model and the error family
surv_a.modelFormula <- cbind(surv, 1 - surv) ~ mass
surv_a.errorFamily <- binomial(link = "logit")

# Sub-set the data frame so only the columns with relevant data are included
surv_a.inputData <- exData[!is.na(exData$surv) & 
                           !is.na(exData$mass) & exData$a==1, ]

```

Next we specify the growth (change in mass) model. We specify a Gaussian regression with body mass at t+1 as the response variable and body mass at t and age class (a) as predictors. 

```{r growthSpec}
# Define the model (error family not needed as assumed to be Gaussian)
growth_f.modelFormula <- mass_next ~ mass
# Sub-set the data frame so only the columns with relevant data are included
growth_f.inputData <- exData[!is.na(exData$mass_next) & 
                            !is.na(exData$mass) & exData$a==0, ]

# Define the model (error family not needed as assumed to be Gaussian)
growth_a.modelFormula <- mass_next ~ mass
# Sub-set the data frame so only the columns with relevant data are included
growth_a.inputData <- exData[!is.na(exData$mass_next) & 
                             !is.na(exData$mass) & exData$a==1, ]

```

Then, we specify two sub-models to describew for fecundity. First, we estimate the probability of a female producing a minimum of one fledging (rep), which is a function of size in the same year. The error is assumed to be binomial with a logit link function. 

```{r repSpec}
# Define the model and the error family
rep.modelFormula <- cbind(repF, 1 - repF) ~ mass_next
rep.errorFamily <- binomial(link = "logit")
# Sub-set the data frame so only the columns with relevant data are included
rep.inputData <- exData[!is.na(exData$repF) & 
                        !is.na(exData$mass_next) & exData$a==1, ]
```

Finally, the second sub-model then estimates the brood size (number of fledgings) per successfully breeding female (fec). Here, we assumed a Poisson error distribution and a log link function is used.

```{r fecSpec}
# Define the model and the error family
fec.modelFormula <- Nfledglings ~ fox_next
fec.errorFamily <- poisson(link = "log")
# Sub-set the data frame so only the columns with relevant data are included
fec.inputData <- exData[!is.na(exData$Nfledglings) &
                        !is.na(exData$mass_next)&exData$a==1,]
```

Because the model fitting mechanism in PaGAn uses [Markov Chain Monte Carlo (MCMC)](https://en.wikipedia.org/wiki/Markov_chain_Monte_Carlo) we can also specify some settings for the fitting algorithm.  This is done by using a list with the following possible entries:
  * __numRuns__ The number of iteration that each chain will run (after the 'burn-in' phase has ended)
* __numChains__ The number of chains to run
* __numBurnIn__ The number of samples at the start of each chain to discard
* __thinDensity__ Thinning parameter for the regression coefficients and error distribution parameters of the model.  This should usually be kept at 1 unless you are needed to run the chains for so long that you are running into memory problems before convergence
* __predictThinDensity__ Thinning parameter for the predictions from the model.  In cases where the dataset is very large then the number of samples for each of the prediction densities can cause memory problems.  This parameter allows for separate control over thinning of the posterior prediction densities whilst maintaining more dense sampling of the regression coefficients

```{r mcmcParams}
# Define the MCMC parameters
mcmcParams <- list(
  numRuns = 1000,
  numChains = 4,
  numBurnIn = 500,
  thinDensity = 1,
  predictThinDensity = 1
)
```

In PaGAn the vital rate models are all fitted simultaneously using the 'ipmNIMBLE' function.  Each vital rate sub-model must be given a name and the naming convention used to pass arguments to the function must be of the format '[sub-model name].[argument]' where '[sub-model name]' is replaced by the name of your vital-rate sub-model and '[argument]' is replaced by an argument to pass to the sub-model.  For example 'surv.modelFormula' would refer to the 'modelFormula' argument of the 'surv' sub-model.  You can name your vital-rate sub-models in any way you like as long as you avoid characters that are not alphanumeric and that you don't start the name of the sub-model with a digit.  You can provide as many sub-models as you like: in this example we are using three but you could have more or fewer as you feel neccessary for your application.  The 'ipmNIMBLE' function understands the following sub-model arguments:
  * __modelFormula__ The formula for the vital rate regression using the standard model notation used in R's [glm function](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/glm)
* __errorFamily__ Specification for the family of the error distribution using R's [family objects](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/family).  If this is not provided then a Gaussian error model is assumed.
  * __inputData__ The data to use for the fitting of the vital rate.  This is optional and global objects will be used instead if this argument is not provided
  * __regCoeffs__ An advanced setting that allows for regularization of the coefficients.  This defaults to "none" but can be "ridge" or "lasso" if those types of regression are required

In addition to the sub-model arguments the 'ipmNIMBLE' has the following optional global arguments:
  * __mcmcParams__ A list object containing the parameters for the MCMC
  * __inputData__ A data frame that will be used as the input data is the 'inputData' arguments is not supplied to a sub-model
  * __numCores__ Defaults to 1 but can set higher if you wish to use multiple cores to fit the model (sometimes useful for large datasets or where there are many vital rate functions).  However, using multicore analysis on Window's systems can result in the status information not being displayed to the console

In this example we therefore call the function using the following arguments;

```{r vitalFit, results='hide', warning=FALSE, message=FALSE}
# Run the IPM
ipmOut <- ipmNIMBLE(mcmcParams = mcmcParams,
  # Define the survival sub-models
  # Fledglings
  surv_f.modelFormula = surv_f.modelFormula, 
  surv_f.errorFamily = surv_f.errorFamily, 
  surv_f.inputData = surv_f.inputData,
  
  # Adults
  surv_a.modelFormula = surv_a.modelFormula, 
  surv_a.errorFamily = surv_a.errorFamily, 
  surv_a.inputData = surv_a.inputData,

  # Define the growth sub-models
  # Fledglings
  growth_f.modelFormula = growth_f.modelFormula, 
  growth_f.inputData = growth_f.inputData,
  #Adults
  growth_a.modelFormula = growth_a.modelFormula, 
  growth_a.inputData = growth_a.inputData,

  # Define the reproductive probability sub-model
  rep.modelFormula = rep.modelFormula, 
  rep.errorFamily = rep.errorFamily, 
  rep.inputData = rep.inputData,
  
  # Define the brood size sub-model
  fec.modelFormula = fec.modelFormula, 
  fec.errorFamily = fec.errorFamily,
  fec.inputData = fec.inputData,
  
  )
```

After the vital rate models have been fit  we can then extract some of the model outputs. The resultant 'ipmOut' object is a list with a named argument for each of the vital rate sub-models. Each of the sub-models has a long list of outputs. For example, for the survival sub-model we can explore the following outputs:

```{r ipmOutput}

# Retrieve the WAIC for the sub-model
ipmOut$surv_a$WAIC

# Retrieve the Bayesian R-squared for the sub-model
setNames(
  c(ipmOut$surv_a$rSquared$mean, 
    ipmOut$surv_a$rSquared$quant0.025, 
    ipmOut$surv_a$rSquared$quant0.975),
  c("expected", "lower95credible", "upper95credible"))

# Plot the DHARMa residual plots for the sub-model
plot(ipmOut$surv_a$DHARMaResiduals)

# Plot marginal posterior distributions of the regression coefficients
ipmOut$surv_f$parameterFigure
ipmOut$surv_a$parameterFigure
ipmOut$growth_a$parameterFigure

# Plot the trace plots of the MCMC of regression coefficients
plot(ipmOut$surv_a$parameterSamples)

# Plot of  Gelman-Rubin convergence diagnostic shrink plots
gelman.plot(ipmOut$surv_a$parameterSamples)

```

## Kernel Creation
The next step in any IPM analysis is to create kernels for life history characteristics that we are interested in.  In PaGAn we define kernels by creating R functions that combines the predictions of the vital rate functions in such a way as to calculate the life history trait that you are interested in.  If your function has any arguments of the form '[sub-model name].[aspect]' where '[sub-model name]' is the name of one of your vital rate sub-models then the relevant 'aspect' of that vital rate is provided to the function.  The current aspects can be requested:
  * __response__ A prediction from the vital rate sub-model (on the response scale) according to an input data set (see later)
* __linpred__ A prediction from the vital rate sub-model (on the linear predictor scale) according to an input data set (see later)
* __sd__ The standard deviation of the error distribution of the sub-model
* __var__ The variance of the error distribution of the sub-model
* __prec__ The precision of the error distribution of the sub-model
* __scale__ The scale of the error distribution of the sub-model (can only be used for error distributions that have a scale parameter)

We next define a data set for which we want to make predictions from. In this example, we want to make predictions of the probabilities of arriving at a mass in year t+1 given a mass in t so we can use the 'seq' function in R to make a sequence of masses that wish to use. This will be used as input data for the vital-rate model predictions ('response' and 'linpred') described earlier.  Our models here all only use 'mass' as a predictor variable so we only need to have mass in the prediction data set. 

However, if our vital rate models had other additional covariates (e.g., temperature or precipitation) then we would need to provide those in our prediction data set too for a given level of that covariate (e.g., mean covariate, 0.5 degrees warming)

```{r testData}
testData <- data.frame(mass = seq(za0_range[1]*0.9, za1_range[2]*1.1, 
                                  length.out = 100),
                       fox_next=0)

testData$mass_next <- testData$mass

# Calculate the distance between each prediction point
cellWidth <- diff(testData$mass[1:2])

```

For example, the growth kernel would be defined by the following kernel function:

```{r growthKern}
# Function to evaluate the growth kernel
growthKern_f <- function(x, growth_f.response, growth_f.sd, cellWidth) {
  
  dnorm(x, growth_f.response, growth_f.sd) * cellWidth
  
}

growthKern_a <- function(x, growth_a.response, growth_a.sd, cellWidth) {
  
  dnorm(x, growth_a.response, growth_a.sd) * cellWidth
  
}
```

The first argument of the kernel function must be the value for which the probability density is being estimated. We can then use the 'ipmKernelEvaluation' function to evaluate this function.

```{r growthKernEval}
# Evaluate the growth kernel
growthMat_f <- ipmKernelEvaluation(ipmOut, 
                                   testData, testData$mass, 
                                   growthKern_f, 
                                   cellWidth = cellWidth)

growthMat_a <- ipmKernelEvaluation(ipmOut, testData, 
                                   testData$mass, 
                                   growthKern_a, 
                                   cellWidth = cellWidth)

# Plot the expected values of the growth kernel matrix - for fledglings
image(testData$mass, testData$mass, t(growthMat_f$expected), 
      xlab = "mass(t)", ylab = "mass(t + 1)")

# Plot the expected values of the growth kernel matrix - for adults
image(testData$mass, testData$mass, t(growthMat_a$expected), 
      xlab = "mass(t)", ylab = "mass(t + 1)")

# Pull out a specific sample from the MCMC samples e.g., 10th (for adults)
image(testData$mass, testData$mass, t(growthMat_a$samples[, , 10]), 
      xlab = "mass(t)", ylab = "mass(t + 1)")

```

You can see from this output that 'ipmKernelEvaluation' function produces a list with two elements: __expected__ which is the expected kernel matrix and __samples__ which is the kernel matrix evaluated for each of the samples from the MCMC.  Because this analysis is Bayesian we do not just have one kernel matrix but whole posterior distribution of kernel matrices.  The user can also specify other arguments to be passed to the kernel function through providing as named arguments in the 'ipmKernelEvaluation' function.

```{r growth/survival}
# Evaluate the growth/survival kernel - fledglings
growthSurvKern_f <- 
  function(x, surv_f.response, growth_f.response, growth_f.sd, cellWidth) {
  surv_f.response * growthKern_f(x, growth_f.response, growth_f.sd, cellWidth)
}
growthSurvMat_f <- ipmKernelEvaluation(ipmOut, 
                                       testData, 
                                       testData$mass, 
                                       growthSurvKern_f, 
                                       cellWidth = cellWidth)

# Plot the expected values of the growth survival kernel matrix
image(testData$mass, testData$mass, t(growthSurvMat_f$expected), 
      xlab = "size(t)", ylab = "size(t + 1)", 
      main="fledgling survival and growth")

# Evaluate the growth/survival kernel - adults
growthSurvKern_a <- 
  function(x, surv_a.response, growth_a.response, growth_a.sd, cellWidth) {
  surv_a.response * growthKern_a(x, growth_a.response, growth_a.sd, cellWidth)
}
growthSurvMat_a <- ipmKernelEvaluation(ipmOut, 
                                       testData, 
                                       testData$mass, 
                                       growthSurvKern_a, 
                                       cellWidth = cellWidth)

# Plot the expected values of the growth survival kernel matrix
image(testData$mass, testData$mass, t(growthSurvMat_a$expected), 
      xlab = "size(t)", ylab = "size(t + 1)", main="adult survival and growth")

```


```{r recszKern}
# Function to evaluate the fledging size kernel
recszKern <- function(x, recsz.response, cellWidth) {
  dnorm(x, recsz.response, recsz.sd) * cellWidth
}
```


```{r recEval}
# Recruitment kernel
recsizeMean <- mean(exData$mass_next,na.rm = T)
recsizeSD <- sd(exData$mass_next,na.rm = T)
sexratio <- 0.5

recKern <- function(x, surv_a.response, rep.response, 
                    fec.response, recsizeMean, recsizeSD, cellWidth) {

  surv_a.response * rep.response * fec.response * 
    sexratio * dnorm(x, recsizeMean, recsizeSD) * cellWidth

}

recMat <- ipmKernelEvaluation(ipmOut, testData, testData$mass, 
                              recKern, recsizeMean = recsizeMean, 
                              recsizeSD = recsizeSD, cellWidth = cellWidth)

# Plot the expected values of the recruitment kernel matrix
image(testData$mass, testData$mass, t(recMat$expected), 
      xlab = "mass(t)", ylab = "mass(t + 1)")

```

Now we build the full kernel by combing the survival, growth and recruitment kernels, using a matrix assembly function. 

```{r FullKern}
# Specify matrices
stgMatrix <- matrix(c(0, 1,
                      1, 1),
                      nrow = 2, ncol = 2, byrow = TRUE)

ipmKernels <- list(growthSurvMat_f, growthSurvMat_a,recMat)
states <- c('Fledgling', 'Adult')
kernelFrom <- c('Fledgling','Adult', 'Adult')
kernelTo <- c('Adult','Adult', 'Fledgling')


FullOut <- megaMatrixAssembly(stgMatrix, ipmKernels, 
                              states, kernelFrom, kernelTo)
```

And now we finally analyse the full kernel using the ipmKernelAnalysis() function and make some plots of the asymptotic growth rate and age class specific stable mass distributions

```{r FullKernAnal}
# Analyse the full kernel (asymptotic growth rate and mass distribution)
fullMatAnalysis <- ipmKernelAnalysis(FullOut)

# Plot the asymptotic growth rate
ggplot(data = data.frame(growthRate = fullMatAnalysis$assymPopGrowth)) + 
  geom_histogram(aes(x = growthRate), fill = "#69b3a2", color = "#e9ecef") +
  geom_vline(xintercept = fullMatAnalysis$assymPopGrowthSummary[4], 
             linetype = "dashed") + 
  geom_vline(xintercept = fullMatAnalysis$assymPopGrowthSummary[5], 
             linetype = "dashed") +
  geom_vline(xintercept = fullMatAnalysis$assymPopGrowthSummary[1]) +
  theme_classic() + ylab("Count") + xlab("Asymptotic Growth Rate")

# Plot the stable mass distribution of fledglings 
p1=ggplot(data = cbind(data.frame(size = testData$mass), 
                       fullMatAnalysis$stableDistSummary[1:100,])) +
  geom_ribbon(aes(x = size, ymin = lower95cred, ymax = upper95cred), 
              fill = "cornflowerblue", color = "#e9ecef") +
  geom_line(aes(x = size, y = mean)) +
  theme_classic() + ylab("") + xlab("Mass") + ggtitle("Fledglings")

# Plot the stable mass distribution of adults 
p2=ggplot(data = cbind(data.frame(size = testData$mass), 
                       fullMatAnalysis$stableDistSummary[101:200,])) +
  geom_ribbon(aes(x = size, ymin = lower95cred, ymax = upper95cred), 
              fill = "#69b3a2", color = "#e9ecef") +
  geom_line(aes(x = size, y = mean)) +
  theme_classic() + ylab("") + xlab("Mass") + ggtitle("Adults")

gridExtra::grid.arrange(p1,p2,nrow=1)

```


``` {r all kernel images}
# Plot expected values of seperate kernels
image(testData$mass, testData$mass, t(growthSurvMat_f$expected), 
      xlab = "mass(t)", ylab = "mass(t + 1)")
image(testData$mass, testData$mass, t(growthSurvMat_a$expected), 
      xlab = "mass(t)", ylab = "mass(t + 1)")
image(testData$mass, testData$mass, t(recMat$expected), 
      xlab = "mass(t)", ylab = "mass(t + 1)")
```
