---
title: "Barnacle goose IPM"
author: "Kate Layton-Matthews"
output:
  pdf_document: default
  html_document:
    df_print: paged
vignette: |
  %\VignetteIndexEntry{Simple IPM Example} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Integral Projection Models for svalbard barnacle geese

```{r libraries, results='hide', warning=FALSE, message=FALSE}
# This chunk will be removed once the package is completed - in the meantime it loads the required libraries for the analysis and imports the relevant source files from the project GitHub pages
library(nimble)
library(parallel)
library(coda)
library(DHARMa)
library(ggplot2)

source("https://raw.githubusercontent.com/joechip90/PaGAn/master/R/mcmcInternals.R")
source("https://raw.githubusercontent.com/joechip90/PaGAn/master/R/glmNIMBLE.R")
source("https://raw.githubusercontent.com/joechip90/PaGAn/master/R/ipmNIMBLE.R")
```

## Goose dataset
Reproductive and survival data from the breeding grounds from 1991-2017.

```{}
# Data loaded seperately: shown here
knitr::kable(head(exData))

str(exData)
za0_range <- round(range(exData$mass[exData$a==0],na.rm=T))
za1_range <- round(range(exData$mass[exData$a==1],na.rm=T))

```

## Vital Rate Regressions

Run vital rate regressions. 
Survival.
```{r survSpec}
# Define the model and the error family
surv.modelFormula <- cbind(surv, 1 - surv) ~ mass * a
surv.errorFamily <- binomial(link = "logit")

# Sub-set the data frame so only the columns with relevant data are included
surv.inputData <- exData[!is.na(exData$surv) & !is.na(exData$mass), ]
```

Growth model. Gaussian regression body mass at t+1 as the response variable and mass at t + age class (a) as predictors.
```{r growthSpec}
# Define the model (error family not needed because it is assumed to be Gaussian if not specified)
growth.modelFormula <- mass_next ~ mass + a
# Sub-set the data frame so only the columns with relevant data are included
growth.inputData <- exData[!is.na(exData$mass_next) & !is.na(exData$mass), ]
```

Sub-model of the probability of producing a fledging (rep). Binomial error distribution and a logit link function is used.
```{r repSpec}
# Define the model and the error family
rep.modelFormula <- cbind(repF, 1 - repF) ~ mass_next
rep.errorFamily <- binomial(link = "logit")
# Sub-set the data frame so only the columns with relevant data are included
rep.inputData <- exData[!is.na(exData$repF) & !is.na(exData$mass_next) & exData$a==1, ]
```

Sub-model for number of fledgings per successful female (fec). Poisson distributed and a log link function is used.
```{r fecSpec}
# Define the model and the error family
fec.modelFormula <- Nfledglings ~ fox_next
fec.errorFamily <- poisson(link = "log")
# Sub-set the data frame so only the columns with relevant data are included
fec.inputData <- exData[!is.na(exData$Nfledglings) & exData$a==1 & !is.na(exData$mass_next), ]
```

Specific mcmc parameters.
```{r mcmcParams}
# Define the MCMC parameters
mcmcParams <- list(
  numRuns = 1000,
  numChains = 4,
  numBurnIn = 500,
  thinDensity = 1,
  predictThinDensity = 1
)
```

Vital rate models are all fitted simultaneously using the 'ipmNIMBLE' function. 
```{r vitalFit, results='hide', warning=FALSE, message=FALSE}
# Run the IPM
ipmOut <- ipmNIMBLE(mcmcParams = mcmcParams,
  # Define the fledgling survival sub-model
  surv.modelFormula = surv.modelFormula, surv.errorFamily = surv.errorFamily, surv.inputData = surv.inputData,

  # Define the fledging growth sub-model
  growth.modelFormula = growth.modelFormula, growth.inputData = growth.inputData,

  # Define the reproductive probability sub-model
  rep.modelFormula = rep.modelFormula, rep.errorFamily = rep.errorFamily, rep.inputData = rep.inputData,
  
  # Define the fecundity sub-model
  fec.modelFormula = fec.modelFormula, fec.errorFamily = fec.errorFamily, fec.inputData = fec.inputData,
  
  )
```

Extract some model outputs, e.g., survival submodel.
```{r ipmOutput}

# Retrieve the WAIC for the sub-model
ipmOut$surv$WAIC

# Retrieve the Bayesian R-squared for the sub-model
setNames(
  c(ipmOut$surv$rSquared$mean, ipmOut$surv$rSquared$quant0.025, ipmOut$surv$rSquared$quant0.975),
  c("expected", "lower95credible", "upper95credible"))

# Plot the DHARMa residual plots for the sub-model
plot(ipmOut$surv$DHARMaResiduals)

# Plot the margianl posterior distributions of the regression coefficients associated with the sub-model
ipmOut$surv$parameterFigure
ipmOut$fec$parameterFigure

# Plot the trace plots of the MCMC of regression coefficients for the sub-model
plot(ipmOut$surv$parameterSamples)

# Plot of the Gelman-Rubin convergence diagnostic shrink plots for the sub-model
gelman.plot(ipmOut$surv$parameterSamples)

```

## Kernel Creation

We next define a data set for which we want to make predictions from (all predictors included in model).
```{r testData}
# Create a prediction data set to evaluate the kernels on
testData <- data.frame(mass = c(seq(za0_range[1], za0_range[2], length.out = 100), 
                                seq(za1_range[1], za1_range[2], length.out = 100)),
                       a = rep(c(0,1),each=100),
                       fox = 0,
                       fox_next = 0)

testData$mass_next <- testData$mass

# Calculate the distance between each prediction point
cellWidth <- diff(testData$mass[1:2])

```

Growth kernel only: 
```{r growthKern}
# Function to evaluate the growth kernel
growthKern <- function(x, growth.response, growth.sd, cellWidth) {
  dnorm(x, growth.response, growth.sd) * cellWidth
}
```

Evaluating and plotting: 
```{r growthKernEval}
# Evaluate the growth kernel
growthMat <- ipmKernelEvaluation(ipmOut, testData, testData$mass, growthKern, cellWidth = cellWidth)

# Plot the expected values of the growth kernel matrix
# Fledglings
image(testData$mass[1:100], testData$mass[1:100], t(growthMat$expected[1:100,1:100]), xlab = "mass(t)", ylab = "mass(t + 1)")

# Adults 
image(testData$mass[101:200], testData$mass[101:200], t(growthMat$expected[101:200,101:200]), xlab = "mass(t)", ylab = "mass(t + 1)")

# Pull out a specific sample from the MCMC samples e.g., 10th 
image(testData$mass[1:100], testData$mass[1:100], t(growthMat$samples[1:100, 1:100 , 10]), xlab = "mass(t)", ylab = "mass(t + 1)")

```

Evaluate growth/survival kernel: 
```{r rest}
# Evaluate the growth/survival kernel
growthSurvKern <- function(x, growth.response, growth.sd, surv.response, cellWidth) {
  surv.response * growthKern(x, growth.response, growth.sd, cellWidth)
}
growthSurvMat <- ipmKernelEvaluation(ipmOut, testData, testData$mass, growthSurvKern, cellWidth = cellWidth)

# Fledglings
image(testData$mass[1:100], testData$mass[1:100], t(growthSurvMat$expected[1:100,1:100]), xlab = "mass(t)", ylab = "mass(t + 1)", main="fledgling survival and growth")

# Adults
image(testData$mass[101:200], testData$mass[101:200], t(growthSurvMat$expected[101:200,101:200]), xlab = "size(t)", ylab = "size(t + 1)", main="adult survival and growth")
```

```{r growthKern}
# Function to evaluate the fledging size kernel
recszKern <- function(x, recsz.response, cellWidth) {
  dnorm(x, recsz.response, recsz.sd) * cellWidth
}
```


```{r rest}
# Recruitment kernel
recsizeMean <- mean(exData$mass_next,na.rm = T)
recsizeSD <- sd(exData$mass_next,na.rm = T)

recKern <- function(x, surv.response, rep.response, fec.response, recsizeMean, recsizeSD, cellWidth) {

  # How to add in if a == 0 then rep = 0???
  
  surv.response * rep.response * fec.response * 0.5 * dnorm(x, recsizeMean, recsizeSD) * cellWidth

}

recMat <- ipmKernelEvaluation(ipmOut, testData, testData$mass, recKern, recsizeMean = recsizeMean, recsizeSD = recsizeSD, cellWidth = cellWidth)

#Fledglings
image(testData$mass[1:100], testData$mass[1:100], t(recMat$expected[1:100,1:100]), xlab = "mass(t)", ylab = "mass(t + 1)")

# Evaluate the full kernel
fullKern <- function(x, growth.response, growth.sd, surv.response, rep.response, fec.response, recsizeMean, recsizeSD, cellWidth) {
  
  growthSurvKern(x, growth.response, growth.sd, surv.response, cellWidth) + 
  
  recKern(x, surv.response, rep.response, fec.response, recsizeMean, recsizeSD, cellWidth) 
  }

fullMat <- ipmKernelEvaluation(ipmOut, testData, testData$mass, fullKern, recsizeMean=recsizeMean, recsizeSD=recsizeSD, cellWidth = cellWidth)

# Fledgings
image(testData$mass[1:100], testData$mass[1:100], t(fullMat$expected[1:100,1:100]), xlab = "mass(t)", ylab = "mass(t + 1)")

# Adults
image(testData$mass[101:200], testData$mass[101:200], t(fullMat$expected[101:200,101:200]), xlab = "mass(t)", ylab = "mass(t + 1)")

# Analyse the full kernel
fullMatAnalysis <- ipmKernelAnalysis(fullMat)

ggplot(data = data.frame(growthRate = fullMatAnalysis$assymPopGrowth)) + geom_histogram(aes(x = growthRate), fill = "#69b3a2", color = "#e9ecef") +
  geom_vline(xintercept = fullMatAnalysis$assymPopGrowthSummary[4], linetype = "dashed") + geom_vline(xintercept = fullMatAnalysis$assymPopGrowthSummary[5], linetype = "dashed") +
  geom_vline(xintercept = fullMatAnalysis$assymPopGrowthSummary[1]) +
  theme_classic() + ylab("Count") + xlab("Asymptotic Growth Rate")

ggplot(data = cbind(data.frame(mass = testData$mass), fullMatAnalysis$stableDistSummary)) +
  geom_ribbon(aes(x = mass, ymin = lower95cred, ymax = upper95cred), fill = "#69b3a2", color = "#e9ecef") +
  geom_line(aes(x = mass, y = mean)) +
  theme_classic() + ylab("") + xlab("mass")

```

